---
title: "Spatial smoothing with BTB R package"
author: "Julien PRAMIL"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial smoothing with BTB R package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 5,
  fig.width = 5,fig.align = 'center'
  
  
)
```


This document will show you :

* How to install 'btb' package
* How to do your first spatial smoothings :
  * densities
  * means
  * rates
  * quantiles smothings

Furthermore, it will introduce a way to map your results using 'mapsf' package and how to save your smoothed spatial data using 'sf'. 

# Install 'btb' package

`btb` is available on CRAN : 

```{r setup , eval = FALSE}
install.packages("btb")
```

But you can also download `btb` from github.com : 

```{r, eval = F}
install.packages("remotes")
remotes::install_github("InseeFrLab/btb")
```


# Let's do it

## Warning with personnal data 

Spatial smoothing generally reduces individual data disclosure. 
However, smoothed data can contain individual information.
Please remain cautious in any case. 

## Smoothing gas station prices

### The data

`btb` package provides several data tables. Lets's use `dfPrix_SP95_2016`.
For every gas station in metropolitan France, this table gives :

  * longitude / latitude coordinates (as numeric variables)
  * annual mean price for unleaded gasoline in € for (2016)

```{r seepoints , eval = T}
library(btb)
data(dfPrix_SP95_2016)
head(dfPrix_SP95_2016)
```

Let's visualize theses stations : 
  
  * First : use `sf` package to transform your data.frame as geometric points.
  * Then : plot it

```{r cartopoints}
library(sf)
sfPrix_SP95_2016 <- st_as_sf(dfPrix_SP95_2016,coords = c("x","y"), crs=2154)
plot(sfPrix_SP95_2016$geometry)
```

## Optional step : from points to aggregate grids

To figure out your spatial distribution before to smooth you data, it can be good to aggregate your points inside a grid (e.g : number of gas stations in 20 km pixels grid). 

`btb` provids the `btb_add_centroids` and the `btb_ptsToGrid` functions to make it easy : 

* First : associate each point with the centroid of its pixel (`btb_add_centroids`)
* Secondly : aggregate your data by centroids
* To finish : associate each centroid coordinates with its geometric polygon (`btb_ptsToGrid`)

```{r, addcentro}
# Step 1 
dfPrix_SP95_2016 <- btb_add_centroids(dfPrix_SP95_2016,iCellSize = 20000,names_coords = c("x","y"))
head(dfPrix_SP95_2016)
```


```{r checkgrid}
# Step 2
library(dplyr)
centro_values <- dfPrix_SP95_2016 %>% group_by(x_centro,y_centro) %>% summarise(pricemean=mean(SP95,rm.na=T))

# Step 3
grid_values <- btb_ptsToGrid(centro_values,sEPSG = 2154,iCellSize = 20000,names_centro = c("x_centro","y_centro"))
nrow(grid_values)
head(grid_values)

```

Once you have your polygons and your aggregated data, you can map it.
Here, we use the 'mapsf' package.

```{r seegrid}
library(mapsf)

mapsf::mf_map(x = grid_values,
       type = "choro",
       var="pricemean",
       breaks = "quantile",
       nbreaks = 5,
       lwd=1,
       leg_val_rnd = 1)
```

This map represents your aggregated (mean price) but not smoothed yet.
Despite its patchwork aspect, this map could be a good first step to better understand your data.

## First smoothing : the density of gas stations

On the example below, we smooth the density of gas stations using 5\~000 km pixels and a 100 km bandwidth. 
Note that we need to create a new dummy variable (equals to 1 for every station). 

```{r smooth_density}

pts_density <- dfPrix_SP95_2016[,c("x","y")]
# Create dummy
pts_density$stations_density <- 1L
head(pts_density)

# Smoothing
smooth_density <- btb_smooth(
  pts = pts_density,
  sEPSG = 2154,
  iBandwidth = 100000,
  iCellSize = 5000)

head(smooth_density)

# Map
mapsf::mf_map(x = smooth_density,
       type = "choro",
       var="stations_density",
       breaks = "quantile",
       nbreaks = 5,
       border = NA,
       leg_val_rnd = 1)
```

Note that `btb_smooth` is *conservative* : 

* Number of gas stations in `pts_density` : `r sum(pts_density$stations_density)`
* Number of gas station in `smooth_density` : `r sum(smooth_density$stations_density)`

## Smoothing means : gas mean price

Smoothing a ratio works almost the same way.
First, you need to smooth both nominator and denominator.
Then, to calculate a properly smoothed ratio, you must calculate the ratio of theses 2 smoothed variables (smoothed nominator / smoothed denominator).
Note that the `btb_smooth` function smoothes all numeric variables in the input points table (parameter `pts`).


```{r smooth_mean_price}
# Prepare your data
pts_meanprice <- dfPrix_SP95_2016[,c("x","y","SP95")]
pts_meanprice$stations_density <- 1L
head(pts_meanprice)

# Smooth both prices and station density
smooth_density <- btb_smooth(
  pts = pts_meanprice,
  sEPSG = 2154,
  iBandwidth = 100000,
  iCellSize = 5000)

head(smooth_density)

# Calculate the smoothed mean (from smoothed nominator and denominator)
smooth_density <- smooth_density %>% mutate(meanprice=SP95/stations_density)
mapsf::mf_map(x = smooth_density,
       type = "choro",
       var="meanprice",
       breaks = "quantile",
       nbreaks = 5,
       border = NA,
       leg_val_rnd = 1)
```

```{r,include=F}
Cstack_info()
```

## Quantile smoothing : smooth the distribution of gas prices

Quantile smoothing is a different methodology. 
Its major benefits are :

  * less sensitive to outliers
  * gives information on the distribution of your data

For more theorical information, please see (XXXXXXXX : formation séquence lissage théorique).



```{r quantile_smooth}

pts_quantiles <- dfPrix_SP95_2016[,c("x","y","SP95")]
head(pts_quantiles)

smooth_quantiles <- btb_smooth(pts = pts_quantiles,sEPSG = 2154,iBandwidth = 100000,iCellSize = 5000,vQuantiles = c(0.5,0.9))

head(smooth_quantiles)

# Median smoothing : 
mapsf::mf_map(x = smooth_quantiles,
       type = "choro",
       var="SP95_05",
       breaks = "quantile",
       nbreaks = 5,
       border = NA,
       leg_val_rnd = 1)

# Smooth the 9th decile :
mapsf::mf_map(x = smooth_quantiles,
       type = "choro",
       var="SP95_09",
       breaks = "quantile",
       nbreaks = 5,
       border = NA,
       leg_val_rnd = 1)

```


## The iNeighbor parameter

Here, we use fiscal data in La Réunion island.
Each point is the centroid of the grid used to publish aggregated data (1\ 000 meters pixels).

Let's smooth the proportion of poors among households with an **automatic grid** (`iNeighbor` parameter absent in `btb_smooth` function). 

In the following example, note that the `btb_smooth` function accepts sf points in input (also the case with `btb_ptsToGrid`). 


```{r ratesmooth}

# Load data
data("reunion")
head(reunion)

# Optional : transform as sf points
sfreunion <- sf::st_as_sf(reunion,coords= c("x","y"), crs = 3727)
plot(sfreunion$geometry)

# btb_smooth with an automatic grid
smooth_reunion <- btb_smooth(sfreunion,iCellSize = 500,iBandwidth = 5000)

# Calculate the ratio
smooth_reunion <- smooth_reunion %>% mutate(prop_poors = 100 * phouhold / houhold)

# map
mapsf::mf_map(x = smooth_reunion,
       type = "choro",
       var="prop_poors",
       breaks = "quantile",
       nbreaks = 5,
       border = NA,
       leg_val_rnd = 1)
```

Now, let's smooth the same ratio, with the same smoothing specifications (`iBandwidth` and `iCellSize`) but with **iNeighbor = 0**. In this case, the automatic grid only uses pixels that contain at least 1 data point (here, at least 1 household). The result is quite different.

```{r neighboors}
smooth_reunion <- btb_smooth(sfreunion,iCellSize = 500,iBandwidth = 5000, iNeighbor = 0)
smooth_reunion <- smooth_reunion %>% mutate(prop_poors = 100 * phouhold / houhold)

mapsf::mf_map(x = smooth_reunion,
       type = "choro",
       var="prop_poors",
       breaks = "quantile",
       nbreaks = 5,
       border = NA,
       leg_val_rnd = 1)
```

## Inspire naming

Using the [Inspire norm](https://inspire.ec.europa.eu/), `btb_smooth` and `btb_ptsToGrid` allow you to name your pixels in a proper way. It could be useful for reuse purpose, merge operations, etc.

You just need to use `inspire = T` : 

```{r inspire}
smooth_reunion <- btb_smooth(sfreunion,iCellSize = 500,iBandwidth = 2000, iNeighbor = 0,inspire = T)
smooth_reunion <- smooth_reunion %>% mutate(prop_poors = 100 * phouhold / houhold)
head(smooth_reunion)

```

Then, to export your geometric data, you can use the `sf::write_sf` function.

```{r export, eval=F}
sf::write_sf("MY/REPOSITORY/myfile.gpkg")
```



## References :

- https://inseefrlab.github.io/formation-r-lissage-spatial/tuto.html
- Formation lissage théorique
- 

## Some interesting usecases 

- https://mobile.twitter.com/DavidZumbach/status/1373166163497213952
- http://r.iresmi.net/2019/05/11/kernel-spatial-smoothing-transforming-points-pattern-to-continuous-coverage/
- https://semba-blog.netlify.app/06/30/2020/kernel-smoothin-of-spatial-data/
- https://mobile.twitter.com/raffverduzco/status/1128075094524350464?lang=bg
- https://githubmemory.com/repo/SNStatComp/awesome-official-statistics-software
